---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# machineryR

**State-machine workflow orchestration for ontology-based processes.**

machineryR models real-world workflows as defined processes with states,
transitions, human-in-the-loop steps, automated triggers, and oversight
dashboards. It sits directly above `actionTypesR` in the stack:

```
actionTypesR   — single-action dispatch
     ↓
machineryR     — multi-step workflow orchestration
     ↓
objectExploreR — process dashboard module (optional)
```

Where `actionTypesR` handles one action at a time, machineryR sequences
multiple actions, enforces ordering, tracks the state each object instance
is in, handles failures and retries, and surfaces what needs human attention.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("CathalByrneGit/machineryR")
```

## Core concepts

| Concept | Description |
|---|---|
| **Process** | A workflow definition: named states + transitions for one object type |
| **Instance** | One object moving through a process; has a current state and history |
| **State** | A named step in the workflow, optionally with an SLA |
| **Transition** | A directed edge from one state to another, fired by a trigger |
| **Trigger** | What fires a transition: an action, a condition, a manual step, or a timeout |

### Trigger types

| Trigger | Constructor | When it fires |
|---|---|---|
| Action | `mac_action_trigger()` | When a named `actionTypesR` action is submitted for this object |
| Condition | `mac_condition_trigger()` | When a SQL expression evaluates to `TRUE` for this object |
| Manual | `mac_manual_trigger()` | When `mac_advance()` is called by an authorised actor |
| Timeout | `mac_timeout_trigger()` | After the instance has been in a state for longer than N hours |

## Quick start

### 1. Define a process

```{r define, eval = FALSE}
library(machineryR)

discharge_process <- mac_process(
  id             = "patient_discharge",
  name           = "Patient Discharge Workflow",
  object_type_id = "Encounter",
  initial_state  = "admitted",
  table_name     = "encounters",   # for condition trigger evaluation
  key_column     = "encounter_id",
  states = list(
    mac_state("admitted",           display_name = "Admitted",           sla_hours = 72),
    mac_state("discharge_eligible", display_name = "Discharge Eligible", sla_hours = 24),
    mac_state("pending_approval",   display_name = "Pending Approval",   sla_hours = 4),
    mac_state("discharged",         display_name = "Discharged",         terminal = TRUE),
    mac_state("escalated",          display_name = "Escalated",          terminal = TRUE)
  ),
  transitions = list(
    mac_transition(
      from    = "admitted",
      to      = "discharge_eligible",
      trigger = mac_condition_trigger("ready_for_discharge = TRUE",
                                      check_interval_mins = 30L)
    ),
    mac_transition(
      from    = "discharge_eligible",
      to      = "pending_approval",
      trigger = mac_action_trigger("InitiateDischarge")
    ),
    mac_transition(
      from    = "pending_approval",
      to      = "discharged",
      trigger = mac_manual_trigger(required_role = "discharge_approver")
    ),
    mac_transition(
      from    = "pending_approval",
      to      = "escalated",
      trigger = mac_timeout_trigger(hours = 4)
    )
  )
)
```

### 2. Create a context and register

```{r register, eval = FALSE}
library(DBI)
library(duckdb)

con <- dbConnect(duckdb(), "workflow.duckdb")
ctx <- mac_context(bundle = "hospital_ops", connection = con)

mac_register(ctx, discharge_process)
#> ✔ Registered process 'patient_discharge' (5 states, 4 transitions).
```

### 3. Start instances

```{r start, eval = FALSE}
iid <- mac_start(ctx, process_id = "patient_discharge", object_key = "enc_001")
#> ✔ Started instance 'inst_...' for object 'enc_001' in process 'patient_discharge'
#>   (state: admitted).

mac_state_of(ctx, iid)
#> [1] "admitted"
```

### 4. Drive transitions

**Condition trigger** — evaluated on a schedule or on demand:

```{r condition, eval = FALSE}
# Mark the encounter as ready in the source table
dbExecute(con, "UPDATE encounters SET ready_for_discharge = TRUE WHERE encounter_id = 'enc_001'")

mac_check_conditions(ctx, process_id = "patient_discharge")
mac_state_of(ctx, iid)
#> [1] "discharge_eligible"
```

**Action trigger** — fired when an `actionTypesR` action is submitted:

```{r action, eval = FALSE}
# Wire machineryR into your actionTypesR context
action_ctx <- mac_wire_actions(ctx, action_ctx)

# Now any submit_action("InitiateDischarge", target_ids = "enc_001") will
# automatically advance the matching instance.

# Or call the trigger directly:
mac_check_action_triggers(ctx,
  action_type_id = "InitiateDischarge",
  target_ids     = "enc_001",
  submission_id  = "sub_abc"
)
mac_state_of(ctx, iid)
#> [1] "pending_approval"
```

**Manual trigger** — a human approves the next step:

```{r manual, eval = FALSE}
mac_advance(ctx, iid, actor = "dr_smith", notes = "Patient stable, ready to go.")
#> ✔ Advanced instance '...': pending_approval -> discharged (by dr_smith).

mac_state_of(ctx, iid)
#> [1] "discharged"
```

**Timeout trigger** — fires automatically after the configured duration:

```{r timeout, eval = FALSE}
# mac_check_timeouts() is called on a schedule (cron, background job, etc.)
mac_check_timeouts(ctx, process_id = "patient_discharge")
# Any instance stuck in pending_approval for > 4 h is moved to escalated.
```

### 5. Inspect history

```{r history, eval = FALSE}
mac_history(ctx, iid)
#>   transition_id instance_id from_state         to_state trigger_type
#> 1       tr_...      inst_...   admitted discharge_eligible    condition
#> 2       tr_...      inst_... discharge_eligible pending_approval   action
#> 3       tr_...      inst_... pending_approval     discharged       manual
```

## Observability

```{r observe, eval = FALSE}
# Counts per state + average time in state + overdue count
mac_summary(ctx, "patient_discharge")
#>                state count avg_hours_in_state overdue_count
#> 1           admitted     3               18.2             0
#> 2 discharge_eligible     1                6.1             1
#> 3   pending_approval     2                5.8             2
#> 4         discharged    12                0.0             0
#> 5          escalated     1                0.0             0

# Which instances are past their SLA?
mac_overdue(ctx, "patient_discharge")

# Which states are the biggest bottlenecks?
mac_bottlenecks(ctx, "patient_discharge")

# All instances currently in a specific state
mac_instances_in(ctx, "patient_discharge", state = "pending_approval")
```

## Guards and callbacks

### Guard functions

A guard blocks a transition when it returns `FALSE`. The transition stays
pending until the guard allows it.

```{r guard, eval = FALSE}
mac_transition(
  from     = "pending_approval",
  to       = "discharged",
  trigger  = mac_manual_trigger(required_role = "discharge_approver"),
  guard_fn = function(instance, ctx) {
    # Block discharge if an audit flag is set
    flag <- DBI::dbGetQuery(ctx$ctx$con,
      "SELECT audit_flag FROM encounters WHERE encounter_id = ?",
      params = list(instance$object_key[[1]]))
    !isTRUE(flag$audit_flag[[1]])
  }
)
```

### `on_enter` callbacks

`on_enter` fires every time an instance enters the target state. Use it to
send notifications, log events, or trigger downstream jobs.

```{r on_enter, eval = FALSE}
mac_transition(
  from     = "pending_approval",
  to       = "discharged",
  trigger  = mac_manual_trigger(),
  on_enter = function(instance, ctx) {
    notify_patient(instance$object_key[[1]])
  }
)
```

## Shiny dashboard

machineryR ships a Shiny module that renders a live process dashboard:

```{r dashboard, eval = FALSE}
library(shiny)

ui <- fluidPage(
  process_dashboard_ui("discharge")
)

server <- function(input, output, session) {
  process_dashboard_server(
    id           = "discharge",
    mac_ctx_r    = reactive(ctx),
    process_id_r = reactive("patient_discharge")
  )
}

shinyApp(ui, server)
```

The dashboard includes:

- **State distribution** bar chart (count per state)
- **Process flow diagram** (states as nodes, transitions as edges, coloured by terminal/active)
- **Overdue instances** table with object key, state, and hours overdue
- **Instance detail** — click a row to see the full transition history

## Validation

`mac_register()` validates the process definition before storing it and
aborts with a clear message if:

- `initial_state` is not among the defined states
- A transition references an undefined `from` or `to` state
- A terminal state has an outgoing transition
- A non-initial, non-terminal state is unreachable from `initial_state`
- A non-terminal state has no outgoing transitions (dead end)
- Duplicate state names are found

```{r validation, eval = FALSE}
bad_process <- mac_process(
  id = "broken", name = "Broken", object_type_id = "Thing",
  initial_state = "start",
  states = list(
    mac_state("start", sla_hours = 1),
    mac_state("end",   terminal  = TRUE),
    mac_state("orphan")             # unreachable — no transition leads here
  ),
  transitions = list(
    mac_transition("start", "end", trigger = mac_manual_trigger()),
    mac_transition("orphan", "end", trigger = mac_manual_trigger())
  )
)

mac_register(ctx, bad_process)
#> Error: Process 'broken': states unreachable from initial_state 'start': orphan
```

## Database schema

machineryR initialises three tables on first use:

| Table | Purpose |
|---|---|
| `mac_processes` | Serialised process definitions |
| `mac_instances` | One row per tracked object instance |
| `mac_transitions` | Full audit log of every state change |

## Package architecture

```
R/
  context.R      mac_context()              schema init, process registry
  process.R      mac_process / mac_state    DSL constructors
                 mac_*_trigger
                 mac_transition
  validation.R   mac_validate_process()     pre-registration checks
  register.R     mac_register()             store + retrieve process defs
                 mac_get_process()
                 mac_list_processes()
  instance.R     mac_start()                instance lifecycle
                 mac_advance()
                 mac_state_of()
                 mac_history()
  engine.R       mac_check_conditions()     automated trigger evaluation
                 mac_check_timeouts()
                 mac_apply_transition()     internal transition workhorse
  integration.R  mac_wire_actions()         actionTypesR bridge
                 mac_check_action_triggers()
  observe.R      mac_summary()              process observability
                 mac_instances_in()
                 mac_bottlenecks()
                 mac_overdue()
  dashboard.R    process_dashboard_ui()     Shiny module
                 process_dashboard_server()
```

## License

MIT
